\documentclass{fast-nuces-bs}

\usepackage{blindtext}
\usepackage{mathptmx}% Times Roman font
\usepackage[T1]{fontenc}
\usepackage{lipsum} 
\usepackage{sectsty}
\usepackage{tikz-uml}
\usepackage{titlesec}
\usepackage{booktabs}
\usepackage{longtable}

% Class will not insert List of Figures; keep command available

% Information about the Thesis 
% -----------------------------------------------------------------------------
% (All are compulsory, do not delete any line other than author Information)
% For e.g., usually there are 3 students in a group, so leave authorone, 
% authortwo, and authorthree, and delete authorfour.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\department{Department of Computer Science}
\faculty{Computer Science}
\degreeyear{2026}
\degreemonth{June}
\degreename{Computer Science}
\campuscity{Islamabad}
\authorone{Umer Farooq}{22I-0891}
\authortwo{Hussan Waseem Syed}{22I-0893}
\authorthree{Muhammad Irtaza Khan}{22I-0911}
\supervisor{Dr. Imran Ashraf}
\sessionduration{2022-2026}
\cosupervisor{Dr. Muhammad Nouman Noor}
\deanname{Dr. Jawwad Ahmed Shamsi}
\directorname{Dr. Waseem Shahzad}
\hodname{Dr. Hasan Mujtaba}
\fypcoordinatorname{Dr. Imran Ashraf}
\title{Q-CANVAS}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Former document starts below this
\begin{document}

% Tighten top-of-page spacing and remove header rule
\renewcommand{\headrulewidth}{0pt}
\setlength{\headsep}{6pt}

% Abstract
\chapter*{Abstract}\thispagestyle{empty}
This proposal presents Q-CANVAS, a unified quantum simulation platform designed to address the critical standardization gap in multi-framework quantum programming. The platform serves as a universal intermediary, translating quantum circuits written in popular frameworks (Qiskit, Cirq, PennyLane) into the standardized OpenQASM 3.0 intermediate representation. Q-CANVAS features a sophisticated web-based IDE with real-time circuit visualization, user authentication, and GitHub-style circuit sharing. The platform integrates with a quantum simulation backend via API to enable hybrid quantum-classical algorithm execution. This project aims to unify the quantum programming experience, accelerating development and learning while fostering collaboration in the quantum computing ecosystem.
\addcontentsline{toc}{chapter}{Abstract}

% Start main matter page numbering from 1 and restore page style
\clearpage
\setcounter{page}{1}
\pagestyle{fancy}
\include{sections/chapter1}
\include{sections/chapter2}
\include{sections/scope}
\include{sections/conclusions}

%\appendix
%\include{appendix1/appendix1}

\bibliographystyle{plain}
\bibliography{bib} 
\addcontentsline{toc}{chapter}{References} 

\end{document}
