\chapter{Conclusion}

This Software Requirements Specification has defined QCanvas as a unified quantum simulation platform centered on OpenQASM 3.0. It established clear project boundaries, detailed functional and non-functional requirements, described the hybrid CPU–QPU execution model, and documented the system’s architecture, interfaces, and data design. Together, these specifications form the contract for development and verification, ensuring the system remains simulation-first, standards-aligned, and focused on educational and research use cases.

\section{Future Work}

The development roadmap for QCanvas continues with Iteration 3 and Iteration 4, focusing on intelligence, extensibility, and deep integration.

\subsection{Iteration 3: Intelligence and Extensibility}
Iteration 3 (Months 5-6) deals with adding educational intelligence and platform extensibility:
\begin{itemize}
    \item \textbf{"Quantum Explain It" Mode:} AI-driven or heuristic explanations of circuit behavior to aid learning.
    \item \textbf{Gamified Learning:} Interactive modules and challenges to teach quantum concepts in an engaging way.
    \item \textbf{Circuit Sharing Platform:} Community features allowing users to publish, share, and fork quantum circuits.
    \item \textbf{Plug-n-Play Language Support:} Refactoring the architecture to support adding new quantum languages with minimal friction.
\end{itemize}

\subsection{Iteration 4: Integration and Insight}
Iteration 4 (Months 7-8) focuses on the final system merger and advanced analytics:
\begin{itemize}
    \item \textbf{Merge QCanvas + QSim:} Full integration of the frontend/orchestrator with the execution engine.
    \item \textbf{API for Execution:} Public API endpoints allowing external developers to run simulations on QCanvas.
    \item \textbf{Advanced Visualization:} Detailed insights including node statistics and resource utilization (qubit/gate usage).
    \item \textbf{User Metrics:} comprehensive analytics on user engagement and circuit complexity to drive platform improvements.
\end{itemize}

This sets the stage for the final release of QCanvas as a comprehensive educational and research tool.
